Building a continuous benchmarking ecosystem in bioinformatics

### 2024-11-18 manifesto-main.tex
mark's edits, before accepting

\author[1,2,*]{Izaskun Mallona}
\author[1,2,3]{Almut Lütge}
\author[2,4]{Charlotte Soneson}
\author[1]{Ben Carrillo}
\author[1,2]{Reto Gerber}
\author[1]{Daniel Incicau}
\author[1,2]{Anthony Sonrel}
\author[1,2,*]{Mark D. Robinson}

\affil[1]{Department of Molecular Life Sciences, University of Zurich, 8057 Zurich, Switzerland}
\affil[2]{SIB Swiss Institute of Bioinformatics, University of Zurich, 8057 Zurich, Switzerland}
\affil[3]{Swiss Data Science Centre, 8092 Zurich, Switzerland}
\affil[4]{Friedrich Miescher Institute for Biomedical Research, 4056 Basel, Switzerland}

@@

Benchmarking (further defined below) is a critical step in method development. Such studies are typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{cao2023-jz}. A benchmarking study can be run by a single person or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

@@

For \emph{method developers}, benchmarking allows their method to be compared to the state of the art, ideally using neutral datasets and metrics to avoid intrinsic bias \cite{Boulesteix2013-vy, Weber2019-el}. A diverse, accessible set of benchmarking datasets and metrics can address this bias and lower the entry barrier for method developers. Moreover, it is important that the new method is compared to the current state of competing methods, to best mimic the choices available to a potential method user. Alternatively, new methods could be compared to previous versions or even different implementations of the `same' method \cite{Rich2024-jh}, to illustrate stability of code changes or improvements made over time. In both cases, the typical approach today would be for the method developer to rerun all the methods and versions to which they want to compare their current method, on all the datasets they wish to use. This can be challenging and time consuming, since not all methods run on the same hardware setup or in the same software environment (e.g., R and Python versions, library dependencies). It also leads to a lot of redundancy across method comparisons (multiple stakeholders implementing similar workflows), which could be reduced by the presence of a benchmarking system where such results would be accessible and extendable. Finally, the results of the benchmark of a new method should be included in the corresponding publication, and thus an easy way of including the new method into the set of results served by the benchmarking system, and generating a snapshot in time for reproducibility, would be of great help to the developer.

@@

Neutrality is a highly desirable property in a benchmark: it means that efforts have been taken to ensure impartiality or unbiasedness (or at least transparency) of the various choices made (e.g., what datasets to include, how exactly to run methods, how to evaluate methods) \cite{Weber2019-el, Jelizarow2010-kr}. One could claim that all MDP benchmarks are non-neutral, since the goal is generally to highlight the virtues of a new method. Similarly, ground truth generation and simulations in MDP benchmarks can unconsciously mirror the method capabilities, mechanisms, or strengths, thus biasing evaluation. Furthermore, there are many field experts who may wish to conduct a benchmark (or participate in a collaborative benchmark) on analyses that involve their own tools, where neutrality is then harder to establish. Here,  recent work to mitigate this conflict involves pre-registration and posting the parameters of the study in advance at a registry \cite{Sullivan2019-wh, Olevska2021-if}. In contrast, BOP benchmarks give, in principle, a higher degree of neutrality, but the challenge to assess neutrality directly remains. Other benchmarking models are possible, from challenges to hackathons, with the hope that a consensus of the masses can further help to establish neutrality. One potential strategy, in the context of the systematic design of benchmarks, is establishing a system where a benchmark definition is parsed into a pre-registration template. Templates exist for simulation studies that could be adapted more generally for benchmarking projects \cite{Siepe2023-tz}.

@@

Here, we discussed the motivation and process of community benchmarking at several levels, from definition to execution, while highlighting strategies and technologies to potentially ease the technical, scientific and contribution challenges.

>>>>

\author[1,2,*]{Izaskun Mallona}
\author[2,3]{Charlotte Soneson}
\author[1]{Ben Carrillo}
\author[1,2,4]{Almut Lütge}
\author[1]{Daniel Incicau}
\author[1,2]{Reto Gerber}
\author[1,2]{Anthony Sonrel}
\author[1,2,*]{Mark D. Robinson}

\affil[1]{Department of Molecular Life Sciences, University of Zurich, 8057 Zurich, Switzerland}
\affil[2]{SIB Swiss Institute of Bioinformatics, University of Zurich, 8057 Zurich, Switzerland}
\affil[3]{Friedrich Miescher Institute for Biomedical Research, 4056 Basel, Switzerland}
\affil[4]{Swiss Data Science Centre, 8092 Zurich, Switzerland}

@@

Benchmarking (defined below) is a critical step in method development. Such studies are typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{cao2023-jz}. A benchmarking study can be run by a single person or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

@@

For \emph{method developers}, benchmarking allows their method to be compared to the state of the art, ideally using neutral datasets and metrics to avoid intrinsic bias \cite{Boulesteix2013-vy, Weber2019-el}; a recent call for reporting checklists was also made \cite{Brooks2024-rq}. A diverse, accessible set of benchmarking datasets and metrics can address this bias and lower the entry barrier for method developers. Moreover, it is important that the new method is compared to the current state of competing methods, to best mimic the choices available to a potential method user. Alternatively, new methods could be compared to previous versions or even different implementations of the `same' method \cite{Rich2024-jh}, to illustrate stability of code changes or improvements made over time. In both cases, the typical approach today would be for the method developer to rerun all the methods and versions to which they want to compare their current method, on all the datasets they wish to use. This can be challenging and time consuming, since not all methods run on the same hardware setup or in the same software environment (e.g., R and Python versions, library dependencies). It also leads to a lot of redundancy across method comparisons (multiple stakeholders implementing similar workflows), which could be reduced by the presence of a benchmarking system where such results would be accessible and extendable. Finally, the results of the benchmark of a new method should be included in the corresponding publication, and thus an easy way of including the new method into the set of results served by the benchmarking system, and generating a snapshot in time for reproducibility, would be of great help to the developer.

@@

Neutrality is a highly desirable property in a benchmark: it means that efforts have been taken to ensure impartiality or unbiasedness (or at least transparency) of the various choices made (e.g., what datasets to include, how exactly to run methods, how to evaluate methods) \cite{Weber2019-el, Jelizarow2010-kr}. One could claim that all MDP benchmarks are non-neutral, since the goal is generally to highlight the virtues of a new method. Similarly, ground truth generation and simulations in MDP benchmarks can unconsciously mirror the method capabilities, mechanisms, or strengths, thus biasing evaluation. Furthermore, there are many field experts who may wish to conduct a benchmark (or participate in a collaborative benchmark) on analyses that involve their own tools, where neutrality is then harder to establish. Here, recent work to mitigate this conflict involves pre-registration and posting the parameters of the study in advance at a registry \cite{Sullivan2019-wh, Olevska2021-if, Teo2024-wk}. In contrast, BOP benchmarks give, in principle, a higher degree of neutrality, but the challenge to assess neutrality directly remains. Other benchmarking models are possible, from challenges to hackathons, with the hope that a consensus of the masses can further help to establish neutrality. One potential strategy, in the context of the systematic design of benchmarks, is establishing a system where a benchmark definition is parsed into a pre-registration template. Templates exist for simulation studies that could be adapted more generally for benchmarking projects \cite{Siepe2023-tz}.

@@

Here, we discussed the motivation and process of community benchmarking at several levels, from definition to execution, while highlighting strategies and technologies to potentially ease the technical and scientific challenges as well as how to organize contributions from communities.
